---
title: "Probability Distributions"
subtitle: "Chapter 3"
author: "Pedro Cardoso-Leite"
format:
  revealjs:
    theme: [default, ../../../assets/revealjs_themes/teaching.scss]
    slide-number: true
    chalkboard: true
    revealjs-plugins:
      - pointer
---



## Why Probability? {.incremental}


- Statistics rely on probability (e.g., how likely are these results under the null hypothesis?)
- SEM tests **how well our model fits the data**.
- **Probability theory** is the engine that allows us to do that.




## Anatomy of a probability distribution

- what are the events that could possibly happen (sample space)?
- how probable is it that a specific event will happen (probability distribution)?

[https://en.wikipedia.org/wiki/Probability_distribution](https://en.wikipedia.org/wiki/Probability_distribution)



## Properties of probabilies

Kolmogorov axioms: 

1. $P(x) \leq 0$
2. $\Sigma_x P(x) = 1$
3. "$\sigma$-additivity": probability of the union of mutually independent events is the sum of probability of each individual event



## 

::: {.columns}

:::: {.column}
Discrete RV

- sums: $\Sigma_x P(x)$)
- $P(x_i)  \leq 1$
::::

:::: {.column}
Continuous RV

- integrals: $\int_x P(x)dx$
- $P(x_i)$ can be larger than 1
::::

::: 




# Exercises

## Example 1: 

Coin flip

- what are the events that could possibly happen?
- how probable is it that a specific event will happen?


## Example 2: 

Throw a six sided die

- what are the events that could possibly happen?
- how probable is it that a specific event will happen?


## Example 3: 

The sum of throwing 2 die

- what are the events that could possibly happen?
- how probable is it that a specific event will happen?


## Example 4: 

The height of a student picked at random

- what are the events that could possibly happen?
- how probable is it that a specific event will happen?


## Example 5: 

The number of emails you will get today

- what are the events that could possibly happen?
- how probable is it that a specific event will happen?








# Famous Distributions

## Which ones do you know? {.center}

##

:::: {.columns}

::: {.column}

"Natural" distributions

- [Uniform distribution](https://en.wikipedia.org/wiki/Continuous_uniform_distribution)
- [Normal distribution](https://en.wikipedia.org/wiki/Normal_distribution)
- [Bernoulli distribution](https://en.wikipedia.org/wiki/Bernoulli_distribution)
- [Beta distribution](https://en.wikipedia.org/wiki/Beta_distribution)
- [Binomial distribution](https://en.wikipedia.org/wiki/Binomial_distribution)
- [Poisson distribution](https://en.wikipedia.org/wiki/Poisson_distribution)
- [Exponential distribution](https://en.wikipedia.org/wiki/Exponential_distribution)

:::

::: {.column}

"Artificial" distributions

- [Student's t-distribution](https://en.wikipedia.org/wiki/Student%27s_t-distribution)
- [Chi-Squared](https://en.wikipedia.org/wiki/Chi-squared_distribution)
- [F-distribution](https://en.wikipedia.org/wiki/F-distribution)

:::

::::


## Exercise 1: <br>match to distributions {.smaller}

1. the rate/percentage of success 
2. the IQ scores
3. the number of emails received in one day
4. the probability of winning the lottery among lottery players
5. the ratio of variances 
6. the distribution of squared errors
7. the number of people cured among the treatment group of 10 people
8. the duration between two students entering MSA
9. a coin landing on heads
10. the distribution of an estiamted mean (unkonwn pop. std.)



















# The 4 types of questions <br>to ask a proability distribution



## The 4 R Functions

For every distribution, R provides:

| Function | What it does | Example |
|----------|-------------|---------|
| `d___()` | Density (height of curve) | `dnorm(1.96)` |
| `p___()` | Cumulative probability (area) | `pnorm(1.96)` |
| `q___()` | Quantile (inverse CDF) | `qnorm(0.975)` |
| `r___()` | Random samples | `rnorm(100)` |



## Exercise: <br>what type of questions are these: {.smaller}

1. pick a random card out of a 52 deck of cards
2. What's the probability of getting a 6 when throwing a die?
3. what IQ would you need to have for 30% of the population to have a lower IQ than you?
4. How do you calculate a p-value?
5. What percentage of people have an IQ larger than 130?
6. what's the proability that a psychology student earns between 5'000€ and 6'000€?
7. guess a number between 1 and 10





## Using Probability Functions

```{r}
#| echo: true
#| eval: true
# Probability of z < 1.96
pnorm(1.96)

# z-score for 97.5th percentile
qnorm(0.975)

# p-value for chi-square = 20, df = 14
1 - pchisq(20, df = 14)

# 10 random numbers from Poisson distribution
rpois(10, 2)


```




## Exercise: simulate IQ scores

- Generate 10 random number from a normal distribution with mean=100 and std=15.
- Calcuate the mean of those IQ numbers. 
- What would you expect?

- Do the same with 100, 1000, and 10000 numbers.
- What changes?





# Why is the Normal distribution used everywhere? {.center}



## The Central Limit Theorem

::: {.incremental}

"The central limit theorem states that under certain (fairly common) conditions, the sum of many random variables will have an approximately normal distribution."

--Wikipeadia

:::


## Exercise: <br>test for yourself that the CLT works!

write code to:

- simulate the throw of n die
- compute the sum of those n die
- store that value and repeat 1000 times
- plot the distribution of those sums 
- does it look "normal"?




## CLT in Action

```{r}
#| echo: false
#| dev: png
#| fig-height: 4
#| fig-width: 9
#| fig-align: center
par(mfrow = c(1, 3))
for (n in c(1, 2, 50)) {
  sms <- replicate(5000, sum(sample(1:6, n, replace = TRUE)))
  hist(sms, breaks = 30, probability = TRUE, col = "steelblue",
       main = paste0("n = ", n, " dice"), xlab = "Sum of dice",
       border = "white")
  curve(dnorm(x, mean(sms), sd(sms)), add = TRUE, col = "red", lwd = 2)
}
```









# Exercises

## See labs



## Questions? {.center}
